---
title: "p8105_hw2_sc5126"
author: "Si Chen"
date: "2023-10-04"
output: github_document
---

**Load library** 
```{r load_library, warning = FALSE, message = FALSE}
library(tidyr)
library(tidyverse)
library(readxl)
library(haven)
```


## Problem 1
**Load dataset** 
Use data in pols-month.csv, unemployment.csv, and snp.csv
```{r load_dataset, warning = FALSE, message = FALSE}
month_data = 
  read_csv(file = "./data/pols-month.csv") %>% 
  janitor::clean_names()

unemployment_data = 
  read_csv(file = "./data/unemployment.csv") %>% 
  janitor::clean_names()

snp_data = 
  read_csv(file = "./data/snp.csv") %>% 
  janitor::clean_names()
```

First, clean the data in pols-month.csv. 
Use separate() to break up the variable mon into integer variables year, month, and day; 
replace month number with month name; 
create a president variable taking values gop and dem, and remove prez_dem and prez_gop; 
and remove the day variable.
```{r 1-1}
tidy_month_data = 
  separate(month_data, col = mon, into=c('year', 'month', 'day'), sep='-', convert = TRUE) %>% 
  mutate(
    month = 
      case_match(
        month, 
        1 ~ "January", 
        2 ~ "February",
        3 ~ "March",
        4 ~ "April",
        5 ~ "May",
        6 ~ "June",
        7 ~ "July",
        8 ~ "August",
        9 ~ "September",
        10 ~ "October",
        11 ~ "November",
        12 ~ "December"),
    month = as.factor(month)) %>% 
  mutate(
    president = recode(prez_gop,
                           "0" ="dem",
                           "1" ="gop",
                           "2" = "gop")
         ) |> 
  select(-prez_dem, -prez_gop, -day)

```

Second, clean the data in snp.csv using a similar process to the above. 
For consistency across datasets, arrange according to year and month, 
and organize so that year and month are the leading columns.
```{r 1-2}
tidy_snp_data = 
  separate(snp_data, col = date, into=c('month', 'day', 'year'), sep='/', convert = TRUE) %>% 
  mutate(
    month = 
      case_match(
        month, 
        1 ~ "January", 
        2 ~ "February",
        3 ~ "March",
        4 ~ "April",
        5 ~ "May",
        6 ~ "June",
        7 ~ "July",
        8 ~ "August",
        9 ~ "September",
        10 ~ "October",
        11 ~ "November",
        12 ~ "December"),
    month = as.factor(month)) %>% 
  select(year, month, close)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. 
This process will involve switching from “wide” to “long” format; 
ensuring that key variables have the same name; 
and ensuring that key variables take the same values.
```{r 1-3}
tidy_unemployment_data = 
  pivot_longer(
    unemployment_data, 
    jan:dec,
    names_to = "month", 
    values_to = "unemployment") %>% 
    mutate(
    month = 
      case_match(
        month, 
        'jan' ~ "January", 
        'feb' ~ "February",
        'mar' ~ "March",
        'apr' ~ "April",
        'may' ~ "May",
        'jun' ~ "June",
        'jul' ~ "July",
        'aug' ~ "August",
        'sep' ~ "September",
        'oct' ~ "October",
        'nov' ~ "November",
        'dec' ~ "December"),
    month = as.factor(month)) %>% 
  select(year,month,unemployment)
```

Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r merge_data}
tidy_snp_data$year = as.integer(paste0('20', tidy_snp_data$year))
result_data = 
  left_join(tidy_month_data, tidy_snp_data) %>% 
  left_join(x=., y = tidy_unemployment_data)
```

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

**Description of the dataset:**\
For pols-month dataset:\
The variables in the pols-month dataset include `r colnames(tidy_month_data)`. 
There are `r nrow(tidy_month_data)` rows and `r ncol(tidy_month_data)` columns in the dataset.
The year of this dataset has a range of (`r range(tidy_month_data$year)`).\

For snp dataset:\
The variables in the snp dataset include `r colnames(tidy_snp_data)`. 
There are `r nrow(tidy_month_data)` rows and `r ncol(tidy_month_data)` columns in the dataset.
The year of this dataset has a range of (`r range(tidy_snp_data$year)`).\

For unemployment dataset:\
The variables in the unemployment dataset include `r colnames(tidy_unemployment_data)`. 
There are `r nrow(tidy_unemployment_data)` rows and `r ncol(tidy_unemployment_data)` columns in the dataset.
The year of this dataset has a range of (`r range(tidy_unemployment_data$year)`).\

For the result dataset:\
The variables in the unemployment dataset include `r colnames(result_data)`. 
There are `r nrow(result_data)` rows and `r ncol(result_data)` columns in the dataset.
The year of this dataset has a range of (`r range(result_data$year)`).\



## Problem 2

```{r problem2_Mr. Trash Wheel}
mr_data = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Mr. Trash Wheel",
             range="A2:N586") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    homes_powered = (weight_tons*500) / 30,
    type = "mr",
    year=as.numeric(year)
  )
  
```

```{r problem2_Professor Trash Wheel}
pt_data = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Professor Trash Wheel",
             range="A2:M108") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    homes_powered = (weight_tons*500) / 30,
    type = "pt",
    year=as.numeric(year)
  )
  
```

```{r problem2_Gwynnda Trash Wheel}
gw_data = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Gwynnda Trash Wheel",
             range="A2:L157") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    homes_powered = (weight_tons*500) / 30,
    type = "gw",
    year=as.numeric(year)
  )
  
```

```{r merge_trash wheel}
merged_trash_wheel = 
  bind_rows(mr_data , pt_data, gw_data)
```


**Description of the dataset:**\
For the merged trash wheel dataset:\
The main variables in the merged trash wheel dataset include dumpster number, collection date, homes powered, and type, which describes a observation's origin. 
There are `r nrow(merged_trash_wheel)` rows and `r ncol(merged_trash_wheel)` columns in the dataset.
The year of this dataset has a range of (`r range(merged_trash_wheel$year)`).\
The total weight of trash collected by Professor Trash Wheel was `r sum(pt_data$weight_tons)`. \
The total number of cigarette butts collected by Gwynnda in July of 2021 was `r filter(gw_data, month == "July", year == 2021, type== "gw") %>%  select(cigarette_butts) %>%  sum()`.



## Problem 3
Import, clean, and tidy the dataset of baseline demographics. 
Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), 
and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 
Discuss important steps in the import process and relevant features of the dataset. 
How many participants were recruited, and of these how many develop MCI? 
What is the average baseline age? What proportion of women in the study are APOE4 carriers?
```{r 3-1}
baseline_data = 
  read_csv("data/MCI_baseline.csv", skip = 1, show_col_types = FALSE) %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, "1" = "Male", "0" = "Female"),
         apoe4 = recode(apoe4, "1"= "carrier", "0" ="non-carrier")) %>% 
  filter(age_at_onset > current_age | age_at_onset == ".")

```

**Discussion:**\
We format the variable names in this dataset and encode sex and APOE4 carrier status into meaningful string values. We filter out those observations with onset age greater than current age and those without a proper onset age since they are less useful. Relevant features include id, current age, age at onset, and education, since they describe the conditions and identification of the subjects.\
There are `r nrow(baseline_data)`participants were recruited and included in the dataset. 
Among them, `r filter(baseline_data, age_at_onset != '.') |> nrow()` developed MCI.\
The average baseline age is `r mean(baseline_data$current_age)`. \
The proportion of the women in the study are APOE4 carriers is `r filter(baseline_data, sex=="Female" & apoe4 =="carrier") |> nrow() / nrow(baseline_data) *100`. 



